Music Recommendation System Using the Million Song Dataset

### 2015-05-08 writeup/MSDgroupwriteup.tex
Update MSDgroupwriteup.tex

\section{Introduction}
With an increasing amount of accessible data, recommendation systems are becoming more viable and important in various industries including retail, entertainment, and even dating. Recommendation systems allow for services to extend their reach to customers by allowing their users to discover things of interest. Hence, there is a strong interest in implementing a good recommendation system so that their user base can grow. It can be said that building such a system is even more important for the music industry because the entire library of artists and songs are far larger compared to something like the movie industry so recommending the right music to a specific user is an even harder task. On the other hand, with the shift from a traditional distribution model utilizing physical CDs or records to digital models where users stream or acquire music from internet services, it is now easier than ever to track user preferences and this presents a new opportunity to better predict and recommend songs to users.

In building a recommendation system, various trade-offs had to be accounted for in the type of models that need to be fit. For example, some methods focus more on the metadata of the users and songs (i.e. content-based, such as the user or song characteristics - age, genre, location, etc. ) while other methods use a collaborative filtering approach where a user feedback system such as a star rating is used without considering other metadata. In this paper, the latter approach is the focus.

@@

The scale of the data was too large to explore on our computers for the algorithms that we used, so we subset the data down to a size easily handled by an average laptop. 
It can be seen in Figure \ref{fig:csongs} that the majority of the songs are listened to by very few users we were able to decrease the number of songs in our subset without losing all of the play counts.

@@

We choose to subset the data to users who have listened to more than 27 songs and songs which were listened to by more than 22 users. This resulted in a dataset which was 10\% the size of the original. The sparsity of the data was .0016, meaning that we only had play counts for 0.16\% of the user-song pairs (i.e. not all users listen to all songs). Despite subsetting the data, though, we are still dealing with $8130$ users and $11861$ songs, so some of our algorithms will take a fair amount of time to run.

@@

Intuitively, the score looks at the percentage of recommendations that were in-line with what the user actually liked using the testing set that were not used by the algorithms during training. This score also takes into account the order of the songs recommended so it is desirable to have the song with the highest confidence to be liked at the beginning of the recommendation list. For example, if a user had a test set list of five songs, recommending these five songs at the very beginning would give a MAP score of 100 while recommending them at the very end of the 500 songs in reverse order would only produce a MAP score of 0.0060. The MAP score can be said to be similar to the AUC of the ROC curve discussed in class without taking into account the order of recommendations.

Two different baseline benchmarks based on \citet{McFee:2012:MSD:2187980.2188222} were used. The first was to simply to recommend the top 500 most popular songs in the training dataset to every user with no user personalization. This naive method produced a MAP score of 0.0138 if the top 500 songs were based on the numbers of users who listened to the song or 0.0126 if the top 500 songs were based on the number of total plays the song had. The second baseline used had the bare amount of user personalization where the most popular songs from the artists that the user had also listened to were ordered first. Although this method was also relatively simple and naive, it resulted in a very significant improvement of the MAP score to produce a value of 0.0448.

@@

\subsection{Probabilistic Matrix Factorization}
Matrix factorization is a type of collaborative filtering algorithm that can be used to build a recommendation system for users based on user feedback of objects (in this case songs). This allows us to recommend songs to users based on their listening history without the need for content based approaches. 
Training and testing matrices were constructed, of which both have $N_1$ users (rows denoted by $u_i$) and $N_2$ songs (columns denoted by $v_j$). As users have only listened to a fraction of all the songs available, the matrices are sparse, containing zeros in all entries except for those in which a user (a user representing a single row) has listened to. The goal of matrix factorization is to factor the training matrix into the product of two matrices $U$ and $V$.  The matrix $U$ is $N_1 \times d$ and the matrix $V$ is $d \times N_2$. Another goal of matrix factorization is typically to learn a low-rank factorization of the original matrix while restricting the patterns that are generated in the product of the factorized matrices (e.g. if a user likes a song in a specific group of songs, the user may also like other songs in that group of songs). The choice of $d$ can be subjective but $20$ is a common value to start with. In the factorized matrices, the predicted rating will be $\hat{M}_{ij} = u_i^\TRANSPOSE  v_j$
Using a coordinate ascent algorithm over 100 iterations, each row ($u_i$) and column ($v_j$) of the training matrix is then updated (equations \ref{eq1} and \ref{eq2}) in order to maximize the log joint likelihood (equation \ref{eq3}).

@@

There are several hyperparameters that can be tweaked in order to achieve a better MAP result using matrix factorization. The main ones that we varied were the variance, the rank $d$ of the factorization, and the number of iterations. As the project was being done using Python and the main computational package (\emph{scikit-learn}) did not have probabilistic matrix factorization functions, we decided to apply the algorithm that we wrote ourselves for a different class \citep{koren2009matrix}.

>>>>

\section{Introduction}
With an increasing amount of accessible data, recommendation systems are becoming more viable and important in various industries including retail, entertainment, and even dating. Recommendation systems allow for services to extend their reach to customers by allowing their users to discover things of interest. Hence, there is a strong interest in implementing a good recommendation system so that their user base can grow. It can be said that building such a system is even more important for the music industry because the entire library of artists and songs are far larger as compared to something like the movie industry so recommending the right music to a specific user is an even harder task. On the other hand, with the shift from a traditional distribution model utilizing physical CDs or records to digital models where users stream or acquire music from internet services, it is now easier than ever to track user preferences and this presents a new opportunity to better predict and recommend songs to users.

In building a recommendation system, various trade-offs had to be accounted for in the type of models that need to be fit. For example, some methods focus more on the metadata of the users and songs (i.e. content-based) such as user or song characteristics (e.g. age, genre, location, etc.) while other methods use a collaborative filtering approach where a user feedback system such as a star rating is used without considering other metadata. In this paper, the latter approach is the focus.

@@

The scale of the data was too large to explore on our computers for the algorithms that we used, so we subset the data down to a size easily handled by an average laptop. 
It can be seen in Figure \ref{fig:csongs} that the majority of the songs were listened to by very few users so we were able to decrease the number of songs in our subset without losing all of the play counts.

@@

We chose to subset the data to users who have listened to more than 27 songs and songs which were listened to by more than 22 users. This resulted in a dataset which was 10\% the size of the original. The sparsity of the data was .0016, meaning that we only had play counts for 0.16\% of the user-song pairs (i.e. not all users listen to all songs). Despite subsetting the data, though, we are still dealing with $8130$ users and $11861$ songs, so some of our algorithms will take a fair amount of time to run.

@@

Intuitively, the score looks at the percentage of recommendations that were in-line with what the user actually liked using the testing set that were not used by the algorithms during training. This score also takes into account the order of the songs recommended so it is desirable to have the song with the highest confidence to be liked at the beginning of the recommendation list. For example, if a user had a test set list of five songs, recommending these five songs at the very beginning would give a MAP score of 1 while recommending them at the very end of the 500 songs in reverse order would only produce a MAP score of 0.0060. The MAP score can be said to be similar to the AUC of the ROC curve discussed in class without taking into account the order of recommendations.

Two different baseline benchmarks based on \citet{McFee:2012:MSD:2187980.2188222} were used. The first was to simply recommend the top 500 most popular songs in the training dataset to every user with no user personalization. This naive method produced a MAP score of 0.0138 if the top 500 songs were based on the numbers of users who listened to the song or 0.0126 if the top 500 songs were based on the number of total plays the song had. The second baseline used had the bare amount of user personalization where the most popular songs from the artists that the user had also listened to were ordered first. Although this method was also relatively simple and naive, it resulted in a very significant improvement of the MAP score to produce a value of 0.0448.

@@

\subsection{Probabilistic Matrix Factorization}
Matrix factorization is a type of collaborative filtering algorithm that can be used to build a recommendation system for users based on user feedback of objects (in this case songs). This allows us to recommend songs to users based on their listening history without the need for content based approaches. 
Training and testing matrices were constructed, of which both have $N_1$ users (rows denoted by $u_i$) and $N_2$ songs (columns denoted by $v_j$). As users have only listened to a fraction of all the songs available, the matrices are sparse, containing zeros in all entries except for those in which a user (representing a single row) has listened to. The goal of matrix factorization is to factor the training matrix into the product of two matrices $U$ and $V$.  The matrix $U$ is $N_1 \times d$ and the matrix $V$ is $d \times N_2$. Another goal of matrix factorization is typically to learn a low-rank factorization of the original matrix while restricting the patterns that are generated in the product of the factorized matrices (e.g. if a user likes a song in a specific group of songs, the user may also like other songs in that group of songs). The choice of $d$ can be subjective but $20$ is a common value to start with. In the factorized matrices, the predicted rating will be $\hat{M}_{ij} = u_i^\TRANSPOSE  v_j$
Using a coordinate ascent algorithm over 100 iterations, each row ($u_i$) and column ($v_j$) of the training matrix is then updated (equations \ref{eq1} and \ref{eq2}) in order to maximize the log joint likelihood (equation \ref{eq3}).

@@

There are several hyperparameters that can be tweaked in order to achieve a better MAP result using matrix factorization. The main ones that we explored were the variance, the rank $d$ of the factorization, and the number of iterations. As the project was being done using Python and the main computational package (\emph{scikit-learn}) did not have probabilistic matrix factorization functions, we decided to apply the algorithm that we wrote ourselves for a different class \citep{koren2009matrix}.
